---
title: "Task1"
author: "Slavka"
date: "10/01/2024"
format:
  html:
    embed-resources: true
editor: source
---

# 11Atomic Types
Question description

Create the following variables related to your chosen object:
- 4 numeric values (e.g., population, temperature, size, rating)  
- 3 integer values (e.g., number of moons, buildings, inhabitants)  
- 2 logical values (e.g., has_water, is_inhabited)  
- 3 character values (e.g., name, region, category)

```{r}
# code for solving task
#integer 
population <- 81702 
area <- 70 #square kilometers 
number_of_districts <- 4 

#charakter 
city_name <- "Presov"
city_old_name <- "Eperjes" 
region <- "Presov Region" 

#numeric
average_temperature <- 9.3 
city_budget_million <- 145.6 
unemployment_rate <- 6.2 
population_density <- 1167.17 

#logical 
has_university <- TRUE 
is_capital <- FALSE
```



# 12Vectors
Create three different vectors:

1. **Numeric vector** combining all your numeric and integer variables.  
   *(Use `c()`.)*

2. **Sequence vector** from any number between 10–50 to 100, Choose step between 2 and 10.  
   *(Use `seq()` or `:`.)*

3. **Logical vector** of length 10 with random `TRUE`/`FALSE` values using `sample()`.

```{r}
#_code for solving task
# 1. Numeric vector
numeric_vector <- c(population, area, number_of_districts, average_temperature,
                    city_budget_million, unemployment_rate, population_density)
numeric_vector


# 2. Sequence vector
sequence_vector <- seq(20, 100, by = 10)
sequence_vector




# 3. Logical vector
logical_vec <- sample(c(TRUE, FALSE), size = 10, replace = TRUE)
logical_vec

```
### 13MATRIX
Create a **3×3 matrix** with random numbers between **-50 and 50**  
(using `sample()` or `runif()`), and name its **rows and columns** according to your theme.

```{r}
#_code for solving task
set.seed(123)  # so results stay the same each time
my_matrix <- matrix(sample(-50:50, 9, replace = TRUE), nrow = 3, ncol = 3)

# Name rows and columns according to your theme
rownames(my_matrix) <- c("Center", "North", "South")
colnames(my_matrix) <- c("Economy", "Education", "Transport")

my_matrix

# Row and column averages
rowMeans(my_matrix)
colMeans(my_matrix)

```
# 14LIST
Create a **list** that combines all your previous objects:  
atomic variables, vectors, and matrix.
Add one extra element — a short **description** of your universe as a string.
```{r}
#_code for solving task
presov_list <- list(
  city_name = city_name,
  region = region,
  population = population,
  numeric_vector = numeric_vector,
  sequence_vector = sequence_vector,
  logical_vec = logical_vec,
  matrix = my_matrix,
  description = "This is the data universe of Presov — a Slovak city full of culture, students, and history!"
)

# View the list
presov_list
```



```{r}

```





```{r}

```





# 15Factor
Create a **factor variable** representing categories within your universe, e.g.:



- `"low"`, `"medium"`, `"high"`  
- `"infant"`, `"child"`, `"adult"`  
- `"poor"`, `"average"`, `"rich"`

Make sure you have **at least 5 observations** total.  
Then use `table()` to show how many objects fall into each category.
```{r}
#code_for solving task

city_size <- factor(c("medium", "large", "large", "medium", "small", "large", "medium"))

# Display counts
table(city_size)
```
# 16Data Frame
Create a **data frame** related to your object that includes:

1. **5–8 observations** (rows)  
2. **At least 3 columns** of different data types  
   (e.g., name, numeric value, logical flag, category)
   
```{r}
#_code for solving task
presov_df <- data.frame(
  District = c("Solivar", "Sekcov", "Trojica", "Calvary", "Delna"),
  Population = c(14000, 21000, 13000, 18000, 15702),
  Has_School = c(TRUE, TRUE, TRUE, FALSE, TRUE),
  Area_km2 = c(12.3, 20.1, 9.4, 15.8, 12.4),
  Category = factor(c("residential", "residential", "historic", "suburban", "mixed"))
)

# Rename columns (optional)
colnames(presov_df) <- c("District_Name", "Population", "Has_School", "Area_km2", "Category")

# Structure and summary
str(presov_df)
summary(presov_df)

# Add new derived column: population density per district
presov_df$Density <- presov_df$Population / presov_df$Area_km2

# Show final data frame
presov_df
```

#Task 2 Yaroslava Bohovych 


## Task 2: Subsetting & Missing Data in Your Mini-Universe

### 2.1. Querying Vectors

Let's start with the vectors you created in **Task 1**.

#2.1.1. From your **Sequence vector** (the one with all your numbers):

- Select and print the 3rd element.
- Select and print elements 2 through 5.
- Select and print all values that are greater than 20 (or another number that makes sense for your data).



```{r}
# Select and print the 3rd element
sequence_vector[3]

# Select and print elements 2 through 5
sequence_vector[2:5]

# Select and print all values greater than 50
sequence_vector[sequence_vector > 50]
```


#2.1.2. From your **Logical vector**:

- Select and print only the TRUE values. (Hint: `your_vector[your_vector]`).
- Use `which()` to get the indices (positions) of all the `TRUE` values.


```{r}
# Select and print only the TRUE values
logical_vec[logical_vec]

# Get indices (positions) of all TRUE values
which(logical_vec)
```


### 2.2. Querying Your "Universe" List

Use your main list from **Task 1**.

- Access and print the **description** string using the `$` operator
- Access and print your **3x3 matrix** using double brackets `[[ ]]`.
- Access and print the **second element** from your **vector element** within the list. (This will be a "list-within-a-list" access).

```{r}
# Access and print the description string using $
presov_list$description

# Access and print your 3x3 matrix using double brackets [[ ]]
presov_list[[7]]   # or presov_list$matrix

# Access and print the 2nd element from your numeric vector inside the list
presov_list$numeric_vector[2]

```


### 2.3. Investigating Your Data Frame

Use the **data frame** you created in **Task 1**.

- Print the **first 3 rows**. (Use `head()`.)
- Print the value from the **4th row, 1st column**.
- Print only the column that contains your logical values (e.g., `is_habitable`) as a vector.

**Logical Subsetting**: Create a new data frame `df_subset` that contains only the rows that meet a specific logical condition.

- Example: `df_subset <- your_df[your_df$population > 1000, ]`

`%in%` **Subsetting**: Create a new data frame `df_subset_2` that selects rows belonging to **two specific categories** from your factor or character column.

- Example: `df_subset_2 <- your_df[your_df$region %in% c("North", "South"), ]`

```{r}
# Print the first 3 rows of your data frame
head(presov_df, 3)

# Print the value from the 4th row, 1st column
presov_df[4, 1]

# Print only the logical column (Has_School)
presov_df$Has_School


# Logical Subsetting: keep only districts with population > 15000
presov_subset <- presov_df[presov_df$Population > 15000, ]
presov_subset

# %in% Subsetting: select districts with "residential" or "historic" category
presov_subset_2 <- presov_df[presov_df$Category %in% c("residential", "historic"), ]
presov_subset_2


```

### 2.4. Handling "Corrupted" Data (`NA`)

Create a copy of your original data frame (e.g., `df_corrupted <- your_df`).

**Introduce 3 `NA`s** into your df_corrupted at specific locations.

- Example: `df_corrupted[2, 3] <- NA`

**Find `NAs`**:

- Use `sum(is.na(df_corrupted)`) to find the **total number** of missing values.
- Use `colSums(is.na(df_corrupted))` to see how many `NA`s are in **each column**.

**Calculate with `NA`s**:

- Calculate the `mean()` of the numeric column where you added an `NA`.

**Filter `NA`s**:

- Use `na.omit(df_corrupted)` to create a new, "clean" data frame that contains only the complete rows.


```{r}
# Create a copy of your original data frame
presov_corrupted <- presov_df

# Introduce 3 NA values at specific locations
presov_corrupted[2, 2] <- NA    # Missing population for Sekcov
presov_corrupted[4, 3] <- NA    # Missing Has_School for Calvary
presov_corrupted[5, 5] <- NA    # Missing Category for Delna

# Find total number of missing values
sum(is.na(presov_corrupted))

# Find how many NAs per column
colSums(is.na(presov_corrupted))

# Calculate mean of Population (will return NA)
mean(presov_corrupted$Population)

# Calculate mean ignoring NAs
mean(presov_corrupted$Population, na.rm = TRUE)

# Remove rows with NAs
presov_clean <- na.omit(presov_corrupted)
presov_clean

```

### 2.5. Bonus Challenge: `which()`

Use `which.max()` or `which.min()` to find the **row number** (index) of the observation with the highest or lowest value in one of your numeric columns.

```{r}
# Find index of the row with the highest Population
highest_pop_index <- which.max(presov_df$Population)

# Print that row
presov_df[highest_pop_index, ]

# Find index of the row with the smallest Area_km2
smallest_area_index <- which.min(presov_df$Area_km2)

# Print that row
presov_df[smallest_area_index, ]
```

## Task 3: Expanding Your Universe

Your "mini-universe" in Tasks 1 & 2 was a simulation. Now, it's time to connect it to the real world.

Your mission is to find a real dataset that expands on your theme, import it into R, and explore it using the dplyr package.

### 3.1. The Quest: Find Your Data

Find one simple, real-world dataset that fits your theme.

- **Where to look:** Kaggle, TidyTuesday, data.gov, or other public data portals.

- **Format:** Any format you are able to read. For simplicity it is proposed to use **textual (csv, txt)** or **Excel** format.

- **Requirements:**

    - At least 50 rows.

    - At least 4 columns.

    - A mix of numeric and character/factor data.

- **Action:**

    1. Download the data file and place it in your R project directory.

    2. In your task file add a line with the link to the data source (e.g., Data from: https://...).
    
    ### 3.2. The Import: Load Your Data

1. **Install** `tidyverse`: If you haven't, run `install.packages("tidyverse")` in your console (just once).

2. **Load libraries:** At the top of your script, add `library(dplyr)` and `library(readr)` (`library(readxl)` in case of Excel file format).

3. **Read the data:** Use `read_csv()` (or `read_excel()`) to load your dataset into a new data frame.

```{r}
# Load Libraries
library(dplyr)
library(readr)

# Load the dataset (make sure zoo.csv is in your project folder)
zoo_data <- read_csv("zoo.csv")

# Inspect the dataset
glimpse(zoo_data)
head(zoo_data)
summary(zoo_data)
```

### 3.3. The Interrogation: Explore with `dplyr`

This is the core of the task. Your goal is to **ask and answer 5 interesting questions** about your new dataset.

You must write the question in plain English (as a comment), followed by the `dplyr` code that answers it.

You **must** use each of the following `dplyr` "verbs" at least once:

- `select()` - To pick specific columns.

- `filter()` - To subset rows based on a condition.

- `arrange()` - To sort your results (e.g., with desc()).

- `mutate()` - To create a new column.

`group_by()` and `summarise()` - To get summary statistics (like `mean()`, `n()`, `max()`, etc.).

**Your 5 Questions (example):**

#*Question 1:
Which 10 animals have the highest number of legs?



```{r}
# Q1: Animals with the most legs
zoo_data %>%
  select(animal_name, legs) %>%
  arrange(desc(legs)) %>%
  head(10)
```


#*Question 2:
How many animals belong to each class type?

```{r}
# Q2: Count animals in each class
zoo_data %>%
  group_by(class_type) %>%
  summarise(count = n())
```

#*Question 3:
Which animals are both predators and airborne?

```{r}
# Q3: Airborne predators
zoo_data %>%
  filter(predator == 1, airborne == 1) %>%
  select(animal_name, predator, airborne)
```

#*Question 4:
Create a new variable: “is_mammal” (1 = hair + milk). List 10 mammals.

```{r}
# Q4: New column for mammals
zoo_mammals <- zoo_data %>%
  mutate(is_mammal = ifelse(hair == 1 & milk == 1, 1, 0)) %>%
  filter(is_mammal == 1) %>%
  select(animal_name, hair, milk)

zoo_mammals %>% head(10)
```

#*Question 5:
Which animals have the fewest legs, sorted alphabetically?

```{r}
# Q5: Animals with the fewest legs
zoo_data %>%
  filter(legs == min(legs)) %>%
  select(animal_name, legs) %>%
  arrange(animal_name)
```




