---
title: "Task1"
author: "Slavka"
date: "10/01/2024"
format:
  html:
    embed-resources: true
editor: source
---

# 11Atomic Types
Question description

Create the following variables related to your chosen object:
- 4 numeric values (e.g., population, temperature, size, rating)  
- 3 integer values (e.g., number of moons, buildings, inhabitants)  
- 2 logical values (e.g., has_water, is_inhabited)  
- 3 character values (e.g., name, region, category)

```{r}
# code for solving task
#integer 
population <- 81702 
area <- 70 #square kilometers 
number_of_districts <- 4 

#charakter 
city_name <- "Presov"
city_old_name <- "Eperjes" 
region <- "Presov Region" 

#numeric
average_temperature <- 9.3 
city_budget_million <- 145.6 
unemployment_rate <- 6.2 
population_density <- 1167.17 

#logical 
has_university <- TRUE 
is_capital <- FALSE
```



# 12Vectors
Create three different vectors:

1. **Numeric vector** combining all your numeric and integer variables.  
   *(Use `c()`.)*

2. **Sequence vector** from any number between 10–50 to 100, Choose step between 2 and 10.  
   *(Use `seq()` or `:`.)*

3. **Logical vector** of length 10 with random `TRUE`/`FALSE` values using `sample()`.

```{r}
#_code for solving task
# 1. Numeric vector
numeric_vector <- c(population, area, number_of_districts, average_temperature,
                    city_budget_million, unemployment_rate, population_density)
numeric_vector


# 2. Sequence vector
sequence_vector <- seq(20, 100, by = 10)
sequence_vector




# 3. Logical vector
logical_vec <- sample(c(TRUE, FALSE), size = 10, replace = TRUE)
logical_vec

```
### 13MATRIX
Create a **3×3 matrix** with random numbers between **-50 and 50**  
(using `sample()` or `runif()`), and name its **rows and columns** according to your theme.

```{r}
#_code for solving task
set.seed(123)  # so results stay the same each time
my_matrix <- matrix(sample(-50:50, 9, replace = TRUE), nrow = 3, ncol = 3)

# Name rows and columns according to your theme
rownames(my_matrix) <- c("Center", "North", "South")
colnames(my_matrix) <- c("Economy", "Education", "Transport")

my_matrix

# Row and column averages
rowMeans(my_matrix)
colMeans(my_matrix)

```
# 14LIST
Create a **list** that combines all your previous objects:  
atomic variables, vectors, and matrix.
Add one extra element — a short **description** of your universe as a string.
```{r}
#_code for solving task
presov_list <- list(
  city_name = city_name,
  region = region,
  population = population,
  numeric_vector = numeric_vector,
  sequence_vector = sequence_vector,
  logical_vec = logical_vec,
  matrix = my_matrix,
  description = "This is the data universe of Presov — a Slovak city full of culture, students, and history!"
)

# View the list
presov_list
```



```{r}

```





```{r}

```





# 15Factor
Create a **factor variable** representing categories within your universe, e.g.:



- `"low"`, `"medium"`, `"high"`  
- `"infant"`, `"child"`, `"adult"`  
- `"poor"`, `"average"`, `"rich"`

Make sure you have **at least 5 observations** total.  
Then use `table()` to show how many objects fall into each category.
```{r}
#code_for solving task

city_size <- factor(c("medium", "large", "large", "medium", "small", "large", "medium"))

# Display counts
table(city_size)
```
# 16Data Frame
Create a **data frame** related to your object that includes:

1. **5–8 observations** (rows)  
2. **At least 3 columns** of different data types  
   (e.g., name, numeric value, logical flag, category)
   
```{r}
#_code for solving task
presov_df <- data.frame(
  District = c("Solivar", "Sekcov", "Trojica", "Calvary", "Delna"),
  Population = c(14000, 21000, 13000, 18000, 15702),
  Has_School = c(TRUE, TRUE, TRUE, FALSE, TRUE),
  Area_km2 = c(12.3, 20.1, 9.4, 15.8, 12.4),
  Category = factor(c("residential", "residential", "historic", "suburban", "mixed"))
)

# Rename columns (optional)
colnames(presov_df) <- c("District_Name", "Population", "Has_School", "Area_km2", "Category")

# Structure and summary
str(presov_df)
summary(presov_df)

# Add new derived column: population density per district
presov_df$Density <- presov_df$Population / presov_df$Area_km2

# Show final data frame
presov_df
```

#Task 2 Yaroslava Bohovych 


## Task 2: Subsetting & Missing Data in Your Mini-Universe

### 2.1. Querying Vectors

Let's start with the vectors you created in **Task 1**.

#2.1.1. From your **Sequence vector** (the one with all your numbers):

- Select and print the 3rd element.
- Select and print elements 2 through 5.
- Select and print all values that are greater than 20 (or another number that makes sense for your data).



```{r}
# Select and print the 3rd element
sequence_vector[3]

# Select and print elements 2 through 5
sequence_vector[2:5]

# Select and print all values greater than 50
sequence_vector[sequence_vector > 50]
```


#2.1.2. From your **Logical vector**:

- Select and print only the TRUE values. (Hint: `your_vector[your_vector]`).
- Use `which()` to get the indices (positions) of all the `TRUE` values.


```{r}
# Select and print only the TRUE values
logical_vec[logical_vec]

# Get indices (positions) of all TRUE values
which(logical_vec)
```


### 2.2. Querying Your "Universe" List

Use your main list from **Task 1**.

- Access and print the **description** string using the `$` operator
- Access and print your **3x3 matrix** using double brackets `[[ ]]`.
- Access and print the **second element** from your **vector element** within the list. (This will be a "list-within-a-list" access).

```{r}
# Access and print the description string using $
presov_list$description

# Access and print your 3x3 matrix using double brackets [[ ]]
presov_list[[7]]   # or presov_list$matrix

# Access and print the 2nd element from your numeric vector inside the list
presov_list$numeric_vector[2]

```


### 2.3. Investigating Your Data Frame

Use the **data frame** you created in **Task 1**.

- Print the **first 3 rows**. (Use `head()`.)
- Print the value from the **4th row, 1st column**.
- Print only the column that contains your logical values (e.g., `is_habitable`) as a vector.

**Logical Subsetting**: Create a new data frame `df_subset` that contains only the rows that meet a specific logical condition.

- Example: `df_subset <- your_df[your_df$population > 1000, ]`

`%in%` **Subsetting**: Create a new data frame `df_subset_2` that selects rows belonging to **two specific categories** from your factor or character column.

- Example: `df_subset_2 <- your_df[your_df$region %in% c("North", "South"), ]`

```{r}
# Print the first 3 rows of your data frame
head(presov_df, 3)

# Print the value from the 4th row, 1st column
presov_df[4, 1]

# Print only the logical column (Has_School)
presov_df$Has_School


# Logical Subsetting: keep only districts with population > 15000
presov_subset <- presov_df[presov_df$Population > 15000, ]
presov_subset

# %in% Subsetting: select districts with "residential" or "historic" category
presov_subset_2 <- presov_df[presov_df$Category %in% c("residential", "historic"), ]
presov_subset_2


```

### 2.4. Handling "Corrupted" Data (`NA`)

Create a copy of your original data frame (e.g., `df_corrupted <- your_df`).

**Introduce 3 `NA`s** into your df_corrupted at specific locations.

- Example: `df_corrupted[2, 3] <- NA`

**Find `NAs`**:

- Use `sum(is.na(df_corrupted)`) to find the **total number** of missing values.
- Use `colSums(is.na(df_corrupted))` to see how many `NA`s are in **each column**.

**Calculate with `NA`s**:

- Calculate the `mean()` of the numeric column where you added an `NA`.

**Filter `NA`s**:

- Use `na.omit(df_corrupted)` to create a new, "clean" data frame that contains only the complete rows.


```{r}
# Create a copy of your original data frame
presov_corrupted <- presov_df

# Introduce 3 NA values at specific locations
presov_corrupted[2, 2] <- NA    # Missing population for Sekcov
presov_corrupted[4, 3] <- NA    # Missing Has_School for Calvary
presov_corrupted[5, 5] <- NA    # Missing Category for Delna

# Find total number of missing values
sum(is.na(presov_corrupted))

# Find how many NAs per column
colSums(is.na(presov_corrupted))

# Calculate mean of Population (will return NA)
mean(presov_corrupted$Population)

# Calculate mean ignoring NAs
mean(presov_corrupted$Population, na.rm = TRUE)

# Remove rows with NAs
presov_clean <- na.omit(presov_corrupted)
presov_clean

```

### 2.5. Bonus Challenge: `which()`

Use `which.max()` or `which.min()` to find the **row number** (index) of the observation with the highest or lowest value in one of your numeric columns.

```{r}
# Find index of the row with the highest Population
highest_pop_index <- which.max(presov_df$Population)

# Print that row
presov_df[highest_pop_index, ]

# Find index of the row with the smallest Area_km2
smallest_area_index <- which.min(presov_df$Area_km2)

# Print that row
presov_df[smallest_area_index, ]
```

## Task 3: Expanding Your Universe

Your "mini-universe" in Tasks 1 & 2 was a simulation. Now, it's time to connect it to the real world.

Your mission is to find a real dataset that expands on your theme, import it into R, and explore it using the dplyr package.

### 3.1. The Quest: Find Your Data

Find one simple, real-world dataset that fits your theme.

- **Where to look:** Kaggle, TidyTuesday, data.gov, or other public data portals.

- **Format:** Any format you are able to read. For simplicity it is proposed to use **textual (csv, txt)** or **Excel** format.

- **Requirements:**

    - At least 50 rows.

    - At least 4 columns.

    - A mix of numeric and character/factor data.

- **Action:**

    1. Download the data file and place it in your R project directory.

    2. In your task file add a line with the link to the data source (e.g., Data from: https://...).
    
    ### 3.2. The Import: Load Your Data

1. **Install** `tidyverse`: If you haven't, run `install.packages("tidyverse")` in your console (just once).

2. **Load libraries:** At the top of your script, add `library(dplyr)` and `library(readr)` (`library(readxl)` in case of Excel file format).

3. **Read the data:** Use `read_csv()` (or `read_excel()`) to load your dataset into a new data frame.

```{r}
# Load Libraries
library(dplyr)
library(readr)

# Load the dataset (make sure zoo.csv is in your project folder)
zoo_data <- read_csv("zoo.csv")

# Inspect the dataset
glimpse(zoo_data)
head(zoo_data)
summary(zoo_data)
```

### 3.3. The Interrogation: Explore with `dplyr`

This is the core of the task. Your goal is to **ask and answer 5 interesting questions** about your new dataset.

You must write the question in plain English (as a comment), followed by the `dplyr` code that answers it.

You **must** use each of the following `dplyr` "verbs" at least once:

- `select()` - To pick specific columns.

- `filter()` - To subset rows based on a condition.

- `arrange()` - To sort your results (e.g., with desc()).

- `mutate()` - To create a new column.

`group_by()` and `summarise()` - To get summary statistics (like `mean()`, `n()`, `max()`, etc.).

**Your 5 Questions (example):**

#*Question 1:
Which 10 animals have the highest number of legs?



```{r}
# Q1: Animals with the most legs
zoo_data %>%
  select(animal_name, legs) %>%
  arrange(desc(legs)) %>%
  head(10)
```


#*Question 2:
How many animals belong to each class type?

```{r}
# Q2: Count animals in each class
zoo_data %>%
  group_by(class_type) %>%
  summarise(count = n())
```

#*Question 3:
Which animals are both predators and airborne?

```{r}
# Q3: Airborne predators
zoo_data %>%
  filter(predator == 1, airborne == 1) %>%
  select(animal_name, predator, airborne)
```

#*Question 4:
Create a new variable: “is_mammal” (1 = hair + milk). List 10 mammals.

```{r}
# Q4: New column for mammals
zoo_mammals <- zoo_data %>%
  mutate(is_mammal = ifelse(hair == 1 & milk == 1, 1, 0)) %>%
  filter(is_mammal == 1) %>%
  select(animal_name, hair, milk)

zoo_mammals %>% head(10)
```

#*Question 5:
Which animals have the fewest legs, sorted alphabetically?

```{r}
# Q5: Animals with the fewest legs
zoo_data %>%
  filter(legs == min(legs)) %>%
  select(animal_name, legs) %>%
  arrange(animal_name)
```


## Task 4. Building a "Toolkit" for Your Universe (Functions)

In Task 3, you performed a "one-off" analysis of your new dataset. In the real world, you'll want to run that analysis again and again.

The best way to do this is by capturing your code inside **functions**. A function is a reusable "tool" that you build to perform a specific job.

Your mission is to build a "toolkit" of functions that can analyse the dataset you found in Task 3.

**Important:** You must use the **same dataset** and theme you used in Task 3.

### 4.1. The "Greeting" Function (A Simple Start)

Let's start with the basics. Write a simple function that prints a welcome message for your "universe".

1. **Write the function:** It should take at least one argument (e.g., `your_name` or `universe_name`).

2. **Inside the function:** Use `print()` or `cat()` to display a message.

3. **Call the function:** After you define the function, you must call it to see the result.

```{r}
welcome_to_my_universe <- function(your_name) {
  cat("---------------------------------------\n")
  cat(" Welcome to the ZOO ANALYSIS TOOLKIT\n")
  cat(" Report prepared by:", your_name, "\n")
  cat("---------------------------------------\n")
}

# Call the function
welcome_to_my_universe("Slavka Bohovych")
```

### 4.2. The "Loading" Function (A Good Practice)

In Task 3, you wrote code to load and inspect your data. Let's turn that into a function. This is a very common and useful practice.

1. **Write the function:** Name it `load_and_inspect`. It should take one argument: `file_path`.

2. **Inside the function:**

  - Load the `readr` and `dplyr` libraries.

  - Use `read_csv(file_path)` to load the data.

  - Print a message (e.g., "Data loaded successfully.").

  - Print the `glimpse()` of the data.

  - Print the `head()` of the data.

  - **Return** the data frame (use the `return()` command).

3. **Call the function:**

```{r}
load_and_inspect <- function(file_path) {
  library(dplyr)
  library(readr)
  
  data <- read_csv(file_path)
  
  cat("\n--- Data Loaded Successfully from:", file_path, "---\n")
  
  print(glimpse(data))
  
  cat("\n--- First 6 Rows ---\n")
  print(head(data))
  
  return(data)
}

# Call:
my_zoo <- load_and_inspect("zoo.csv")
```

### 4.3. The "Query" Function (The Core of the Task)

This is the most important step. You will turn one of your `dplyr` queries from **Task 3** into a reusable function.

Look back at your Task 3 code. You probably wrote something like this:

  - `my_data |> filter(population > 1000000)`

  - `my_data |> filter(planet_type == "Gas Giant")`

That 1000000 or "Gas Giant" is a "magic number" or "magic string". A function lets you turn it into a parameter.

1. **Choose a query:** Pick one of your `filter()` or `summarise()` queries from Task 3.

2. **Write the function:**
  
  - Give it a descriptive name (e.g., `find_pollution_by_disctrict`, `filter_cities_by_population`).

  - It must take at least two arguments: **data** (for the data frame) and a **parameter** for the value you want to change (e.g., `pop_threshold` or `type_name`).

  - The function should perform the `dplyr` query and `return` the result.


```{r}
find_animals_with_traits <- function(data, airborne_value, predator_value) {
  result <- data %>%
    filter(airborne == airborne_value, predator == predator_value) %>%
    select(animal_name, airborne, predator)
  
  return(result)
}

# Example call:
find_animals_with_traits(my_zoo, airborne_value = 1, predator_value = 1)
```

### 4.4. The "Reporting" Function

Let's combine everything. Write a single function that calculates several key statistics for a specific **group** or **category** in your data.

1. **Write the function:** Name it `generate_report`. It should take `data` and a `category_name` as arguments (e.g., a specific city, planet type, or region).

2. **Inside the function**:

    - `filter()` your data for only the `category_name` provided.

    - Calculate 3-4 interesting summary statistics (e.g., `mean()`, `max()`, `min()`, `n()`).

    - Use `cat()` to print a nicely formatted "report" to the console.

**Example:**

```{r}
generate_report <- function(data, class_number) {

  class_data <- data %>% 
    filter(class_type == class_number)
  
  avg_legs <- mean(class_data$legs, na.rm = TRUE)
  predator_ratio <- mean(class_data$predator, na.rm = TRUE)
  count_animals <- nrow(class_data)
  
  cat("----------------------------------------\n")
  cat(" REPORT FOR CLASS TYPE:", class_number, "\n")
  cat("----------------------------------------\n")
  cat("Number of Animals:", count_animals, "\n")
  cat("Average Number of Legs:", round(avg_legs, 2), "\n")
  cat("Proportion of Predators:", round(predator_ratio, 2), "\n")
  cat("----------------------------------------\n\n")
}

# Run reports for two classes
generate_report(my_zoo, 1)
generate_report(my_zoo, 6)
```
## Task 5. Visualising Your Universe (Base R & ggplot2)

You have found your data (Task 3) and built tools to analyse it (Task 4). Now, you must present your findings to customers.

Your goal is to create four specific visualizations based on the real-world dataset you imported in Task 3.

**Prerequisites:**

  - Use the dataset you found and cleaned in Task 3.

  - Make sure `library(ggplot2)` is loaded.

  - **Crucial:** Every plot must have a **Title**, **x-axis label**, and **y-axis label** that make sense for your specific theme.
  
### 5.1. The Quick Sketch (Base R Histogram)

Sometimes you need a quick look at the distribution of your data without loading heavy packages.

  1. Choose one **numeric variable** from your dataset (e.g., `price`, `temperature`, `speed`).

  2. Use the `hist()` function from base R.

  3. Customise the `main` title, `xlab`, and the bar `col` (colour).
```{r}
# Histogram of number of legs
hist(zoo_data$legs,
     main = "Distribution of Number of Legs in Zoo Animals",
     xlab = "Number of Legs",
     col = "lightblue",
     border = "white")
```

### 5.2. The Relationship Map (ggplot2 Scatter Plot)

Now, let's look for patterns using the industry-standard package, **ggplot2**.

  1. Choose **two numeric variables** to see if they are correlated.

  2. Create a scatter plot using `geom_point()`.
  
  3. Add `geom_smooth()` to add a relationship between variables to the plot.

  4. Add `labs()` to rename your axes and title.
```{r}
library(ggplot2)

ggplot(data = zoo_data, aes(x = legs, y = eggs)) +
  geom_point(alpha = 0.6, size = 3, colour = "blue") +
  geom_smooth(colour = "gray") +
  labs(
    title = "Relationship Between Number of Legs and Egg Laying",
    subtitle = "Zoo Animals Dataset",
    x = "Number of Legs",
    y = "Eggs (1 = lays eggs, 0 = does not)",
    color = "Legend"
  ) +
  theme_minimal()
```

### 5.3. The Comparison (ggplot2 Boxplot)

3. The Comparison (ggplot2 Boxplot)

Boxplots are excellent for comparing groups. Let's see how different categories in your universe perform.

  1. Choose a **categorical (or logical) variable** for the X-axis (e.g., city, planet_type, species).

  2. Choose a **numeric variable** for the Y-axis (e.g., price, gravity, lifespan).

  3. Create a plot using `geom_boxplot()`.

  4. **Bonus:** Use `fill` to colour the boxes by the same category.

```{r}
ggplot(data = zoo_data, aes(x = factor(class_type), y = legs, fill = factor(class_type))) +
  geom_boxplot() +
  labs(
    title = "Comparison of Number of Legs Across Animal Classes",
    x = "Class Type",
    y = "Number of Legs"
  ) +
  theme_light() +
  theme(legend.position = "none")
```

### 5.4. The "Deep Dive" (Faceting or Custom)

The most powerful feature of `ggplot2` is faceting—creating small multiples of the same plot for different groups.

  1. Create a plot (histogram, bar plot, or scatter plot).

  2. Use `facet_wrap(~ your_category_variable)` to split the plot into a separate panel for each group in your data.

  3. This is incredibly useful for comparing distinct groups (e.g., separate charts for each City or Year).
```{r}
ggplot(data = zoo_data, aes(x = legs)) +
  geom_histogram(bins = 10, fill = "coral", color = "black") +
  facet_wrap(~ class_type) +
  labs(
    title = "Distribution of Animal Leg Counts by Class Type",
    x = "Number of Legs",
    y = "Count"
  ) +
  theme_minimal()
```

### 5.5. Final Step: Exporting (Saving Your Work)

A plot is useless if you can't put it in a report.

  1. Choose your favourite plot from the exercises above.

  2. Run the code to display it in R.

  3. Immediately run the following command to save it to your project folder:

```{r}
ggsave("my_universe_plot.png", width = 8, height = 6)
```

